RationalGRL: A Framework for Argumentation and Goal Modeling

### 2017-11-01 formalframework.tex
formalframework cleanup & new example

In Section~\ref{sect:gmas} we developed a list of critical questions and argument schemes by analyzing transcripts of discussions about the development of a traffic simulator. The resulting list is shown in Table~\ref{table:argument-schemes}. We then presented the RationalGRL framework in secton~\ref{sect:overview}, consisting of a modeling language, a metamodel, and various examples of using the argument schemes and critical questions. In section~\ref{sect:meth+tool} we explained how this framework can be used by practitioners by explaining the methodology and tool support.

In this section we present a formalization of RationalGRL based on formal logic. This is done for multiple reasons: (i) Most approaches in formal argumentation use formal logic, allowing us to employ existing technique directly in order to compute which arguments are accepted and which are rejected, (ii) we can be more precise about how critical questions are answered, (iii) we can show that RationalGRL models can be translated in to valid GRL models and visa versa in a precise way, and (iv) the formal approach is a basis for automating the framework in terms of tool support.

In the first two subsections we formalize a static representation of our framework: We provide a formal specification of a GRL model based on the GRL metamodel (Section \ref{sect:metamodel}) in the first subsection, and we extend this with arguments and attack links in the second subsection, hereby obtaining a formal specification of a RationalGRL model.

In the third subsection we develop algorithms in order to translate a GRL model into a RationalGRL model, and visa versa.

In the fourth and final subsection we turn to the dynamics of our framework. We develop algorithms for instantiating argument schemes and for answering critical questions.

@@

In this subsection we formalize a GRL model based on the GRL metamodel and the jUCMNav implementation. We first formalize elements of GRL (intentional elements and actors), and then formalize the links.

\subsubsection{Intentional Element, Beliefs and Actors}

Throughout this section, we adopt the convention that variables start with a lowercase letter (e.g, $id, i, j, name, ie, goal$), and sets and constants start with an uppercase letter (e.g., $Type, AND, Goal$). We start by defining a few general elements of our language that we use in subsequent definitions.

\begin{definition}[General definitions]
\label{def:set-definitions}
We define the following sets:
\begin{itemize}
\item $IETypes = \{Softgoal, Goal, Task, Resource\}$,
\item $LinkTypes = \{PosContr, NegContr, Dep, Decomp\}$\footnote{Recall that for decomposition links we only distinguish between positive and negative contribution. Extending the formalization model to include more values is straightforward.},
\item $Types = IETypes \cup LinkTypes\cup\{Actor, GenArg, ActIE\}$,
\item $Names$ is a finite set of strings,
\item $DecompTypes = \{AND,OR,XOR\}$.
\end{itemize}
\end{definition}
Next, we define an intentional element.

@@

The definition above is sufficient to capture all intentional elements used in GRL. %In the next definition, we present some simplifying notational conventions.
%\begin{definition}[Notation]
%\label{def:notation}
%We adopt three conventions simplifying our notation:
%\begin{itemize}
%\item
%We refer to the element of a tuple using the dot (".") notation. That is, we may for instance refer to the id, type, name and decomposition type of an $IE$ with respectively $ie.id$, $ie.type$, $ie.name$, and $ie.decomptype$.
%\item 
%We refer to a set of elements with the same id $i$  using the $i$ subscript on the set. For instance a set of IEs with id $i$ is denoted by $IE_i$, and if this is a single element, we denote it by $ie_i$\footnote{In GRL, there always exists at most one element for every id (see Def.~\ref{def:grl-model}), in RationalGRL, however, this condition does not hold (see Def.~\ref{def:rationalgrl-model}).} For instance, we may refer to the intentional element $ie = (0, Goal, \text{Make profit}, AND)$ with $ie_0$ and write $ie_0.type = Goal$, $ie_0.name = \text{Make profit}$, and $ie_0.decompositiontype = AND$. 
%\item
%We can also refer to intentional elements of a specific type simply by $type_{id}$. For instance, we can abbreviate the element in the previous item with $goal_0$ and write $goal_0.name = \text{Make profit}$, and $goal_0.decomptype = AND$.
%\end{itemize}
%\end{definition}
Throughout this section we use the example in Figure~\ref{fig:example-small}. The IEs in this example can be formalized using Def.~\ref{def:ie} as follows:
\begin{itemize}
\item (1, Resource, \text{Car objects}, AND),
\item (2, Softgoal, \text{Realistic simulation}, AND),
\item (3, Softgoal, \text{Simple design}, AND),
\item (4, Goal, \text{Generate traffic}, XOR),
\item (5, Task, \text{Create new cars}, AND),
\item (6, Task, \text{Keep same cars}, AND).
\end{itemize}
%Using the short-hand notation defined in Def.~\ref{def:notation}, we can make the following statements:
%\begin{itemize}
%\item $goal_4.name = \text{Generate traffic}$
%\item $goal_4.decomptype = XOR$
%\item $resource_1.name = \text{Car objects}$
%\end{itemize}
%\begin{figure}[ht]
%\centering
%\includegraphics[width=\columnwidth]{img/Example1-new.pdf}
%\caption{Example GRL model (reprint of Fig.~\ref{fig:example-small})}
%\label{fig:example-small2}
%\end{figure} 
We now define actors.

\begin{definition}[Actor]
\label{def:actor}
An actor is a tuple $act=(id,type, name)$, where:
\begin{itemize}
\item $id\in\mathbb{N}$ is the identifier of the actor, 
\item $type = Actor$ states that this tuple is an actor.
\item $name\in Names$ is a string description of its name.
\end{itemize}
A set of actors is denoted by $Act$.
\end{definition}
An actor is simply a pair consisting of its identifier and its name. Similar to intentional elements, we may refer to $act = (id,name)$ with $act_{id}$ and write $act_{id}.name$ to refer to its name. We can formalize the actor of Figure~\ref{fig:example-small} as $act=(0,Actor,\text{Traffic Simulator})$.%, and we can for instance state $act_0.name = \text{Traffic Simulator}$.

@@

\begin{definition}[Actor-IE Relations]
\label{def:act-ie-relation}
An \emph{Actor-IE relation} is a tuple $(i, j)$, where:
\begin{itemize}
%\item $id\in\mathbb{N}$ is the identifier of the relation, 
%\item $ActIE$ states that this tuple is an Actor-IE relation,
\item $i\in\mathbb{N}$ is an id of an actor,
\item $j\in\mathbb{N}$ is an id of an IE.
\end{itemize}

@@

Note that Actor-IE relations do not have an $id$, as they signify relations between elements of a GRL model (i.e., intentional elements and actors) and are themselves not elements of a GRL model. The relation between the \emph{Traffic Simulator} actor ($id = 0$) and the \emph{Car objects} resource, for example, can be formalized as $(0, 1)$.

\subsubsection{Links}

@@

\begin{definition}[GRL Link]
\label{def:link}
A \emph{GRL link} is a tuple $link = (id,type,src,dest)$, where:
\begin{itemize}
\item $id\in \mathbb{N}$ is the unique identifier of the link,
\item $type\in LinkTypes$ specifies the type of the link, 
\item $src\in \mathbb{N}$ is the identifier of the IE from which the link starts,
\item $dest\in \mathbb{N}$ is the identifier of the IE at which the link ends.
\end{itemize}

@@

Similar to IEs, links have identifiers as well. In Fig.~\ref{fig:example-small}, there are five links:

\begin{itemize}
\item $(7, Dep, 4, 1)$
\item $(8, PosContr, 5, 2)$
\item $(9, NegContr, 5, 3)$
\item $(10, Decomp, 5, 4)$
\item $(11, Decomp, 6, 4)$
\end{itemize}

@@

\subsubsection{GRL models}

@@

The definition of a GRL model collects all the previously defined tuples into a single definition. For completeness, we now provide the full specification of Figure~\ref{fig:example-small}. This model is formalized as $GRL=(IE, Act, R_{ActIE}, Link)$, where:
\begin{flalign*}
&IE=&\{&(1, Resource, \text{Car objects}, AND),&\\
&   &  &(2, Softgoal, \text{Realistic simulation}, AND),&\\
&   &  &(3, Softgoal, \text{Simple design}, AND),&\\
&   &  &(4, Goal, \text{Generate traffic}, XOR),&\\
&   &  &(5, Task, \text{Create new cars}, AND),&\\
&   &  &(6, Task, \text{Keep same cars}, AND)\}&\\
&Act=& &\{(0, Actor, \text{Traffic Simulator})\}&\\
&R_{ActIE}=& &\{(0, 1), (0,2), (0,3), (0, 4), (0,5),(0,6)\}&\\
&Link=&\{&(7, Dep, 4, 1)&\\
&     & &(8, PosContr, 5, 2)&\\
&     & &(9, NegContr, 5, 3)&\\
&     & &(10, Decomp, 5, 4)&\\
&     & &(11, Decomp, 6, 4)\}&\\
\end{flalign*}

The definition of a GRL model (Def.~\ref{def:grl-model}) only sums up the elements of the model and not the constraints that make a \emph{valid} GRL model, viz.

@@

Let us briefly verify that our previous formalization of Figure~\ref{fig:example-small} satisfies all the constraints of Def.~\ref{def:valid-grl-model}:
\begin{enumerate}
\item All elements in the formalization have different ids, so this constraint is satisfied.
\item $R_{ActIE}$ contains one element for each IE with id $i$, so this constraint is satisfied as well. Note that links are not related to actors. This is in line with the GRL metamodel.
\item Since we have only one actor with id 0, and this is the only actor that appears in $R_{ActIE}$, this constraint is satisfied.
\item All links connect IEs: the contribution links connect elements with ids 2, 3, and 5, which are all IEs; the decomposition links connect elements with ids 4, 5, and 6, which are all IEs; and the dependency link connects id 1 with 4, which are both IEs.
\end{enumerate}

@@

In order to develop a logical framework for RationalGRL, we extend the GRL logical framework of the previous section by adding two elements (see Figure~\ref{fig:rationalgrllegend}):
\begin{itemize}
\item A new element called \emph{generic argument},
\item A new link called \emph{attack link}
\end{itemize}

\begin{figure}[ht]
\centering
\includegraphics[width=\columnwidth]{img/Example1-new-attack.pdf}
\caption{Example RationalGRL model (extension of Fig.~\ref{fig:example-small2})}
\label{fig:example-small3}
\end{figure}

We illustrate the new elements using Figure~\ref{fig:example-small3}, which is an extension of Figure~\ref{fig:example-small2}.

We start with the new element which we call the \emph{Generic Argument}.

\begin{definition}[Generic Argument]
\label{def:generic-argument}
A generic argument $ga \in \mathbb{N}\times \{GenArg\}\times Names$ is a tuple such that $ga=(id, GenArg, name)$ means
\begin{itemize}
\item $id\in \mathbb{N}$ is the identifier of the generic argument,
\item $GenArg$ states that the tuple is a generic argument,
\item $name\in Names$ is a string description of its name.
\end{itemize}
We may refer to the argument with id $i$ simply with $ga_i$. A set of arguments is denoted by $GA$.
\end{definition}

A generic counterargument is simply an argument that can be used to attack any previous argument, or an IE. Note that one of the constraints of a GRL model (Def.~\ref{def:grl-model}) is that GRL links (Def.~\ref{def:link}) should connect IEs, which means that in GRL generic arguments cannot be connected with GRL links. This is desired, since generic arguments are not part of GRL, so they should also not be connected using GRL links.

In Fig.~\ref{fig:example-small3}, there are two generic counterargument which can be formalized as $(18,GenArg, \text{Redundant})$ and $(19, GenArg, \text{Necessary})$.

@@

\paragraph{Rationale and example} We define an argument as either a generic argument or any of the GRL elements or links. This captures the specification in the RationalGRL metamodel (Fig.~\ref{fig:metamodel}) in which the class \textsf{Argument} is a superclass of \textsf{GenericArgument} and \textsf{GRLModelElement}, which again is a superclass of both \textsf{ElementLink} and \textsf{GRLLinkableElement}. A \textsf{GRLLinkableElement} is in turn a superclass of \textsf{Actor} and \textsf{IntentionalElement}. In sum, we define an argument simply as any one of the GRL elements or links, or a generic argument.

In Fig.~\ref{fig:example-small3}, the actor, all IEs, all links, and all generic counterarguments are arguments. We will give a full formalization of this after the definition of a RationalGRL model (Def.~\ref{def:rationalgrl-model}).

\begin{definition}[Attack Link]
\label{def:link:attack}
Given a set of arguments $Arg$, an attack link $att\in Arg\times Arg$ is a relation such that $att=(A_i,A_j)$ means:
\begin{itemize}
\item $A_i\in Arg$ is the argument performing that attack,
\item $A_j\in Arg$ is the argument being attacked.
\end{itemize}
Intuitively, $att=(A_i,A_j)$ means that argument $A_i$ is attacking argument $A_j$. A set of attack links is denoted by $Att$.
\end{definition}

\paragraph{Rationale and example} The attack link is the only link that RationalGRL adds to GRL. In Fig~\ref{fig:example-small3}, there are three attack links. The first two attack links originate from generic counterarguments and involve the following three arguments:
\begin{itemize}
\item $A_1 = (1, Resource, \text{Car objects}, AND)$
\item $A_{18} = (18, GenArg, \text{Redundant})$
\item $A_{19} = (19, GenArg, \text{Necessary})$
\end{itemize}

These two attack links are formalized as $(A_{18},A_{1})$ and $(A_{19}, A_{18})$.

The third attack link is an attack link created by \emph{replacing} and existing argument (we will discuss this process in more detail in Sect.~\ref{sect:algorithms}). It involves two version of the same IE:
\begin{itemize}
\item $A_{4} = ((4, Goal, \text{Generate traffic}, AND), Goal)$
\item $A_{20} = ((4, Goal, \text{Generate traffic}, XOR), Goal)$
\end{itemize}

This attack link is formalized at $(A_{20},A_4)$.

The last example shows an important difference between RationalGRL models and valid GRL models: While a valid GRL model disallows multiple elements with the same identifier (Def.~\ref{def:valid-grl-model}, condition 1), RationalGRL models do not enforce this restriction. This is because it is possible to create multiple arguments for the same element, where argument contains different content for the same element. This is what the example above also shows. However, the set of \emph{accepted} elements in a RationalGRL should all have unique identifier (see Def.~\ref{def:valid-rationalgrl-model}).

@@

\paragraph{Rationale and example} The definition of a RationalGRL model collects all the previously defined tuples into a single definition. For completeness, we now provide the full specification of Figure~\ref{fig:example-small3}. Let us first enumerate all the arguments used in this example:
\begin{flalign*}
A_0 = &(0, Actor, \text{Traffic simulator})&\\
A_1 = &(1, Task, \text{Car objects}, AND),&\\
A_2 = &(2, Softgoal, \text{Real. sim.}, AND),&\\
A_3 = &(3, Softgoal, \text{Simple des.}, AND),&\\
A_4 = &(4, Goal, \text{Generate traffic}, AND),&\\
A_5 = &(5, Task, \text{Create new cars}, AND),&\\
A_6 = &(6, Task, \text{Keep same cars}, AND),&\\
A_7 = &(7, Dep, 4, 1),&\\
A_8 = &(8, PosContr, 5, 2),&\\
A_9 = &(9, NegContr, 5, 3),&\\
A_{10} = &(10, Decomp, 5, 4),&\\
A_{11} = &(11, Decomp, 6, 4),&\\
A_i = &(i, ActIE, 0,i+11),ActIE), \text{for } 1\le i\le 6&\\
A_{18} = &(18, GenArg, \text{Redundant}),&\\
A_{19} = &(19, GenArg, \text{Necessary}),&\\
A_{20} = &(4, Goal, \text{Generate traffic}, XOR),&\\
\end{flalign*}

This model is then formalized as $RationalGRL=(Arg, Att)$ where:
\begin{flalign*}
Arg = &\{A_0,A_1,\ldots,A_{20}\}\\
Att=&\{(A_{18},A_1), (A_{19},A_{18}), (A_{20},A_{4})\}
\end{flalign*}

All the arguments and the attack relations of this RationalGRL model are shown in Figure~\ref{fig:example-small4}. Note the arguments for Actor-IE containment (arguments $A_{12}$ to $A_{17}$) have been omitted from this figure for readability. It can be read from the figure that two arguments are currently rejected, namely $A_4$ and $A_{19}$. However, we did not yet make precise how exactly this is computed. We will do so in the following definitions.

\begin{figure}[ht]
\centering
\includegraphics[width=\columnwidth]{img/Example1-new-arguments}
\caption{Argumentation framework of RationalGRL model from Fig.~\ref{fig:example-small3}}
\label{fig:example-small4}
\end{figure}

In order to compute when an argument is accepted and when not we use argumentation semantics.  We use the standard approach here, which is known as \emph{Dung's semantics}. The following notions are preliminary.

\begin{definition}[Argumentation Framework~\cite{Dung1995}]
\label{def:argumentation-framework}
An \emph{argumentation framework} $AF=(Arg,Att)$ consists of a set of arguments $Arg$ and a set of attack relations $Att\subseteq Arg\times Arg.$
\end{definition}

We see that the definition of an argumentation framework is very close to that of a RationalGRL model. This allows us to use the following results directly.

\begin{definition}[Attack, conflict-freeness, defense, and admissibility~\cite{Dung1995}] Suppose an argumentation framework $AF=(Arg,Att)$, two sets of arguments $S\cup S'\subseteq Arg$, and some argument $A\in Arg$. We say that
\begin{itemize}
\item $S$ \emph{attacks} $A$ if some argument in $S$ attacks $A$,
\item $S$ \emph{attacks} $S'$ if some argument in $S$ attacks some argument in $S'$,
\item $S$ is \emph{conflict-free} if it does not attack itself,
\item $S$ \emph{defends} $A$ if for each $B$ such that $B$ attacks $A$, $S$ attacks $B$,
\item $S$ is \emph{admissible} if $S$ is conflict-free and defends each argument in it.
\end{itemize}
\end{definition}

\begin{figure}[ht!]
\centering
\begin{tikzpicture}
        \node[minimum size=1cm] (a19) [argNodeIN] at (-3,0) {$A_{19}$};
        \node[minimum size=1cm] (a18) [argNodeIN] at (0,0) {$A_{18}$};
        \node[minimum size=1cm] (a1) [argNodeIN] at (3,0) {$A_1$};

         \path
    (a19) edge [attackLink] (a18)
    (a18) edge [attackLink] (a1);

\end{tikzpicture}
\caption{Example argumentation framework, subset of Fig.~\ref{fig:example-small4}.}
\label{fig:goalmodeling:arg2}
\end{figure}

Let us explain these definitions using the example argumentation framework in Figure~\ref{fig:goalmodeling:arg2}, which is a subset of Figure~\ref{fig:example-small4} containing only arguments $A_1, A_{18}$, and $A_{19}$. In this example, there are two admissible sets: $\{A_{19}\}$ and $\{A_1, A_{19}\}$. In the last admissible set, $A_{19}$ defends $A_1$ against its attacker $A_{18}$. Sets containing both $A_{18}$ and either $A_1$ or $A_{19}$ are not conflict free, and the sets $\{A_1\}$ and $\{A_{18}\}$ do not defend themselves against $A_{18}$ and $A_{19}$, respectively.

>>>>

In Section~\ref{sect:overview}, we have shown through a language definition and informal examples from our case study that it is possible to trace elements of the goal model back to their underlying arguments (\textbf{requirement 2}), and that it is possible to determine the effect of changes in the underlying argumentation on the goal model, and vice versa (\textbf{requirement 3}). A formal rendition of this traceability will be presented in this section, in which we present a logical formalization of RationalGRL. This is done for multiple reasons: (i) Most approaches in formal argumentation (cf. Section~\ref{sect:background:pras}) use formal logic, allowing us to employ existing technique directly in order to compute which arguments are accepted and which are rejected, (ii) we can be more precise about how critical questions are answered, (iii) we can show that RationalGRL models can be translated in to valid GRL models and visa versa in a precise way, and (iv) the formal approach is a basis for automating the framework in terms of tool support.

In Sections \ref{sect:formalframework:grl} and \ref{sect:formalframework:rationalgrl} we formalize a static representation of our framework based on the metamodel Figure~\ref{fig:metamodel}). We first provide a formal specification of a GRL model in Section~\ref{sect:formalframework:grl}, and we extend this with arguments and attack links in the second subsection, hereby obtaining a formal specification of a RationalGRL model. In Section~\ref{sect:formalframework:translation} we then present algorithms in order to translate a GRL model into a RationalGRL model, and vice versa. Finally, in Section~\ref{sect:algorithms} we turn to the dynamics of our framework, developing algorithms for instantiating argument schemes and answering critical questions and thus formally capturing the \textsf{INTRO}, \textsf{REPLACE} and \textsf{DISABLE} operations of which examples were previously given in Section~\ref{sect:overview:examples}.

@@

We formalize a GRL model based on the metamodel (Figure~\ref{fig:metamodel}), starting with intentional elements and actors.

Throughout this section, we adopt the convention that variables start with a lowercase letter (e.g, $id$, $i$, $j$, $name$, $goal$), and sets and constants start with an uppercase letter (e.g., $Type, AND, Goal$). We start by defining general elements of our language that we use in subsequent definitions.

\begin{definition}[General definitions]
\label{def:set-definitions}
We define the following sets:
\begin{itemize}
\item $IETypes = \{Softgoal, Goal, Task, Resource\}$,
\item $LinkTypes = \{PosContr, NegContr, Dep, Decomp\}$\footnote{Recall that for decomposition links we only distinguish between positive and negative contribution. Extending the formalization model to include more values is straightforward.},
\item $Types = IETypes \cup LinkTypes\cup\{Actor, ActIE, GenArg\}$,
\item $Names$ is a finite set of strings,
\item $DecompTypes = \{AND,OR,XOR\}$.
\end{itemize}
\end{definition}
Next, we define an intentional element.

@@

The definition above is sufficient to capture all intentional elements (IEs) used in GRL. %In the next definition, we present some simplifying notational conventions.
%\begin{definition}[Notation]
%\label{def:notation}
%We adopt three conventions simplifying our notation:
%\begin{itemize}
%\item
%We refer to the element of a tuple using the dot (".") notation. That is, we may for instance refer to the id, type, name and decomposition type of an $IE$ with respectively $ie.id$, $ie.type$, $ie.name$, and $ie.decomptype$.
%\item 
%We refer to a set of elements with the same id $i$  using the $i$ subscript on the set. For instance a set of IEs with id $i$ is denoted by $IE_i$, and if this is a single element, we denote it by $ie_i$\footnote{In GRL, there always exists at most one element for every id (see Def.~\ref{def:grl-model}), in RationalGRL, however, this condition does not hold (see Def.~\ref{def:rationalgrl-model}).} For instance, we may refer to the intentional element $ie = (0, Goal, \text{Make profit}, AND)$ with $ie_0$ and write $ie_0.type = Goal$, $ie_0.name = \text{Make profit}$, and $ie_0.decompositiontype = AND$. 
%\item
%We can also refer to intentional elements of a specific type simply by $type_{id}$. For instance, we can abbreviate the element in the previous item with $goal_0$ and write $goal_0.name = \text{Make profit}$, and $goal_0.decomptype = AND$.
%\end{itemize}
%\end{definition}
Throughout this section we use the example in Figure~\ref{fig:example-small}. This example contains various IEs which can be formalized using Definition~\ref{def:ie}, for example, 
$(1, Resource, \text{Car objects}, AND)$, $(2, Softgoal,$ Realistic simulation$, AND)$ and $(5, Goal,$ Generate traffic$, XOR)$.

%Using the short-hand notation defined in Def.~\ref{def:notation}, we can make the following statements:
%\begin{itemize}
%\item $goal_4.name = \text{Generate traffic}$
%\item $goal_4.decomptype = XOR$
%\item $resource_1.name = \text{Car objects}$
%\end{itemize}
%\begin{figure}[ht]
%\centering
%\includegraphics[width=\columnwidth]{img/Example1-new.pdf}
%\caption{Example GRL model (reprint of Fig.~\ref{fig:example-small})}
%\label{fig:example-small2}
%\end{figure} 
We now define actors.

\begin{definition}[Actor]
\label{def:actor}
An actor is a tuple $act=(id,type, name)$, where:
\begin{itemize}
\item $id\in\mathbb{N}$ is the identifier of the actor, 
\item $type = Actor$ states that this tuple is an actor.
\item $name\in Names$ is a string description of its name.
\end{itemize}
A set of actors is denoted by $Act$.
\end{definition}
We can now formalize the actor in our example (Figure~\ref{fig:example-small}) as $(0,Actor,\text{Traffic Simulator})$.%, and we can for instance state $act_0.name = \text{Traffic Simulator}$.

@@

\begin{definition}[Actor-IE Relations]
\label{def:act-ie-relation}
An \emph{Actor-IE relation} is a tuple $(type, i, j)$, where:
\begin{itemize}
%\item $id\in\mathbb{N}$ is the identifier of the relation, 
\item $type = ActIE$ states that this tuple is an Actor-IE relation,
\item $i\in\mathbb{N}$ is an id of an actor,
\item $j\in\mathbb{N}$ is an id of an IE.
\end{itemize}

@@

Note that Actor-IE relations do not have an $id$: they are themselves not elements of a GRL model but rather indicate relations between elements of a GRL model (i.e., intentional elements and actors). The relation between the \emph{Traffic Simulator} actor ($id = 0$) and the \emph{Car objects} resource ($id=1$), for example, can be formalized as $(0, 1)$.

@@

\begin{definition}[GRL Link]
\label{def:link}
A \emph{GRL link} is a tuple $link = (id,type,src,dest)$, where:
\begin{itemize}
\item $id\in \mathbb{N}$ is the unique identifier of the link,
\item $type\in LinkTypes$ specifies the type of the link, 
\item $src\in \mathbb{N}$ is the identifier of the source IE,
\item $dest\in \mathbb{N}$ is the identifier of the destination IE.
\end{itemize}

@@

Similar to IEs, links have identifiers as well. Some of the links in our example (Figure~\ref{fig:example-small}) are $(8, Dep, 4, 1)$, $(9, PosContr, 5, 2)$ and $(11, Decomp, 5, 4)$.

@@

We can now form GRL models as follows.

@@

The full specification of Figure~\ref{fig:example-small} is as follows.

\begin{flalign*}
&IE=&\{&(1, Resource, \text{Car objects}, AND),&\\
&   &  &(2, Softgoal, \text{Realistic simulation}, AND),&\\
&   &  &(3, Softgoal, \text{Simple design}, AND),&\\
&   &  &(4, Softgoal, \text{Easy to use}, AND),&\\
&   &  &(5, Goal, \text{Generate traffic}, XOR),&\\
&   &  &(6, Task, \text{Create new cars}, AND),&\\
&   &  &(7, Task, \text{Keep same cars}, AND)\}&\\
&Act=& &\{(0, Actor, \text{Traffic Simulator})\}&\\
&R_{ActIE}=& &\{(ActIE, 0, 1), \ldots, (ActIE,0,7)\}&\\
&Link=&\{&(8, Dep, 4, 1)&\\
&     & &(9, PosContr, 6, 2)&\\
&     & &(10, NegContr, 6, 3)&\\
&     & &(11, Decomp, 6, 5)&\\
&     & &(12, Decomp, 7, 5)\}&\\
\end{flalign*}

Definition \ref{def:grl-model} only sums up the elements of the model and not the constraints that make a \emph{valid} GRL model, viz.

@@

Let us briefly verify that our previous formalization of Figure~\ref{fig:example-small} satisfies all the constraints of Definition~\ref{def:valid-grl-model}:
\begin{enumerate}
\item All elements in the formalization have different ids, so this constraint is satisfied.
\item $R_{ActIE}$ contains one element for each IE with id $i$, so this constraint is satisfied as well. Note that, in line with the GRL metamodel, links are not related to actors. 
\item There is one actor ($id=0$) and this is the only actor that appears in $R_{ActIE}$ so this constraint is satisfied.
\item All links connect IEs: the contribution links connect elements with ids 2, 3, and 5, which are all IEs; the decomposition links connect elements with ids 4, 5, and 6, which are all IEs; and the dependency link connects id 1 with 4, which are both IEs.
\end{enumerate}

@@

In order to develop a logical framework for RationalGRL, we extend the GRL logical framework of the previous section by adding two elements (see Figure~\ref{fig:rationalgrllegend}), namely \emph{generic arguments} and \emph{attack links}. We illustrate the new elements using the RationalGRL model in Figure~\ref{fig:example-small3}, which is an extension of Figure~\ref{fig:example-small}.

\begin{figure}[b]
\centering
\includegraphics[width=\columnwidth]{img/Example1-new-attack.pdf}
\caption{Example RationalGRL model (extension of Fig.~\ref{fig:example-small})}
\label{fig:example-small3}
\end{figure}

\begin{definition}[Generic Argument]
\label{def:generic-argument}
A generic argument is a tuple $ga=(id, type, name)$, where:
\begin{itemize}
\item $id\in \mathbb{N}$ is the identifier of the generic argument,
\item $type = GenArg$ states that the tuple is a generic argument,
\item $name\in Names$ is a string description of its name.
\end{itemize}
The set of generic arguments is denoted by $AE$.
\end{definition}

So a generic argument is any element in the RationalGRL model that is not an intentional element or an actor. In the example, some of the generic arguments are $(13, GenArg, \text{Redundant})$ and $(14,$ $GenArg,$ Necessary$)$. Note that a constraint of a GRL model (Definition~\ref{def:grl-model}) is that GRL links should connect IEs, which means that in generic arguments cannot be connected with GRL links.

We can now define \emph{arguments}.

@@

This definition captures the specification in the RationalGRL metamodel (Figure~\ref{fig:metamodel}) in which the class \textsf{Argument} is a superclass of \textsf{GenericArgument} and \textsf{GRLModelElement}. In sum, we define an argument simply as any one of the GRL elements or links, or a generic argument.

Examples of arguments in Figure~\ref{fig:example-small3} are $A_5 = (5,$ $Goal,$ Generate traffic$, AND)$, $A'_5 = (5, Goal,$ Generate traffic$, XOR)$, $A_4 = (4, Softgoal,$ Easy to use$, AND)$ and $A_{13}=(13,$ $ArgElem,$ Redundant$)$. Note that the two arguments $A_5$ and $A'_5$ show an important difference between RationalGRL models and valid GRL models: While a valid GRL model disallows multiple elements with the same identifier (Definition~\ref{def:valid-grl-model}, condition 1), RationalGRL models do not enforce this restriction. This is because it is possible to create multiple arguments for the same element, where each argument contains different content for the same element. However, the set of \emph{accepted} elements in a RationalGRL should all have unique identifier (see Definition~\ref{def:valid-rationalgrl-model}).

\begin{definition}[Attack Link]
\label{def:link:attack}
Given a set of arguments $Arg$, an attack link is a relation such that $att=(A_i,A_j)$ means:
\begin{itemize}
\item $A_i\in Arg$ is the argument performing that attack,
\item $A_j\in Arg$ is the argument being attacked.
\end{itemize}
The set of attack links is denoted by $Att$.
\end{definition}

As an example, take the arguments $A_5$ for the goal (Generate traffic, $AND$), $A'_5$ for the goal (Generate traffic, $XOR$), $A_4$ for the softgoal (Easy to use) and the generic argument $A_{13}$ (Redundant) against the softgoal (Easy to use). Given these arguments there are two attack links (Figure~\ref{fig:example-small3}), namely $(A'_{5},A_{5})$ and $(A_{13}, A_{4})$.

We now define RationalGRL models.

@@

The definition of a RationalGRL model collects all the previously defined GRL and RationalGRL elements in a single definition. For completeness, we now provide the full specification of Figure~\ref{fig:example-small3}. Let us first enumerate all the arguments used in this example:

\begin{flalign*}
A_0 = &(0, Actor, \text{Traffic simulator})&\\
A_1 = &(1, Task, \text{Car objects}, AND),&\\
A_2 = &(2, Softgoal, \text{Realistic simulation}, AND),&\\
A_3 = &(3, Softgoal, \text{Simple design}, AND),&\\
A_4 = &(4, Softgoal, \text{Easy to use}, AND),&\\
A_5 = &(5, Goal, \text{Generate traffic}, AND),&\\
A'_5 = &(5, Goal, \text{Generate traffic}, XOR),&\\
A_6 = &(6, Task, \text{Wrap cars}, AND),&\\
A'_6 = &(6, Task, \text{Create new cars}, AND),&\\
A_7 = &(7, Task, \text{Keep same cars}, AND),&\\
A_8 = &(8, Dep, 4, 1),&\\
A_9 = &(9, PosContr, 6, 2),&\\
A_{10} = &(10, NegContr, 6, 3),&\\
A_{11} = &(11, Decomp, 6, 5),&\\
A_{12} = &(12, Decomp, 7, 5),&\\
A_{13} = &(13, GenArg, \text{Redundant}),&\\
A_{14} = &(14, GenArg, \text{Necessary}),&\\
A_{15} = &(15, GenArg, \text{No effect}),&\\
A_{16} = &(16, GenArg, \text{Changes traffic flow}),&\\
A_{17} = &(ActIE,0,1),\ldots,A_{23}=(ActIE,0,7)&\\
\end{flalign*}
This model is then formalized as $RatGRL=(Arg, Att)$ where:
\begin{flalign*}
Arg = &\{A_0,\ldots,A_5, A'_5,A_6,A'_6,A_7,\ldots,A_{23}\}\\
Att= &\{(A'_{5},A_5), (A'_{6},A_{6}), (A_{13},A_{4}), (A_{14},A_{13}),\\
     & (A_{15},A_{9}),(A_{16},A_{15})\}\\
\end{flalign*}

%All the arguments and the attack relations of this RationalGRL model are shown in Figure~\ref{fig:example-small4}. Note the arguments for Actor-IE containment (arguments $A_{12}$ to $A_{17}$) have been omitted from this figure for readability. It can be read from the figure that two arguments are currently rejected, namely $A_4$ and $A_{19}$. However, we did not yet make precise how exactly this is computed. We will do so in the following definitions.

%\begin{figure}[ht]
%\centering
%\includegraphics[width=\columnwidth]{img/Example1-new-arguments}
%\caption{Argumentation framework of RationalGRL model from Fig.~\ref{fig:example-small3}}
%\label{fig:example-small4}
%\end{figure}

From Figure~\ref{fig:example-small3}, it can be read from that arguments $A_5$, $A_6$, $A_{13}$ and $A_{15}$ are currently disabled (rejected). However, we did not yet make precise how exactly this is computed. We will do so in the following definitions.
\begin{figure}[b]
\centering
\includegraphics[]{img/Example1-new-arguments.pdf}
\caption{Example argumentation framework, subset of the RationalGRL model from Figure~\ref{fig:example-small3}.}
\label{fig:goalmodeling:arg2}
\end{figure}
In order to compute when an argument is accepted and when not we use argumentation semantics.  We use the standard approach by Dung~\cite{Dung1995}.

\begin{definition}[Argumentation Framework]
\label{def:argumentation-framework}
An \emph{argumentation framework} $AF=(Arg,Att)$ consists of a set of arguments $Arg$ and a set of attack relations $Att\subseteq Arg\times Arg.$
\end{definition}

Note that the definition of a RationalGRL model (Definition~\ref{def:rationalgrl-model}) is exactly the same as the definition for an argumentation framework. This allows us to use the following definitions directly.

\begin{definition}[Attack, conflict-freeness, defense, admissibility, preferred extension] \label{def:semantics}Suppose an argumentation framework $AF=(Arg,Att)$, two sets of arguments $S\cup S'\subseteq Arg$, and some argument $A\in Arg$. We say that
\begin{itemize}
\item $S$ \emph{attacks} $A$ iff some argument in $S$ attacks $A$,
\item $S$ \emph{attacks} $S'$ iff some argument in $S$ attacks some argument in $S'$,
\item $S$ is \emph{conflict-free} iff it does not attack itself,
\item $S$ \emph{defends} $A$ iff for each $B$ such that $B$ attacks $A$, $S$ attacks $B$,
\item $S$ is \emph{admissible} iff $S$ is conflict-free and defends each argument in it.
\item $S$ is a \emph{preferred extension} iff it is a maximal (w.r.t. set inclusion) admissible set.
\end{itemize}
\end{definition}
